Modelling the Retinotectal projection - can competition provide a stopping mechanism?

### 2021-11-22 paper/paper.tex
Got to a point where I need to create some more figures

\section{Model}

Our model was inspired by the agent-based model of \cite{simpson_simple_2011}
and, in common with that work, consists of agents representing axonal growth
cones, originating from a square retina and moving upon a square tectum of
side length 1 in arbitrary units.
%
\citet{simpson_simple_2011} model three contributions to the movement of the growth
cones; chemoaffinity, competition for space and axon-axon interactions. The
first two of these are modelled without reference to a particular mechanism;
each agent is provided with globally-supervised information about its
instantaneous location and its target location, allowing it to move along a
vector towards the target. Competition is implemented as a movement of
constant distance along the vector between two agents if those two agents are
within a threshold distance of one another. Interaction is signal-induced
repulsion, which is triggered if two agents are both within the threshold
distance of each other and have compatible EphA receptor expression.

We sought to replace the globally-supervised mechanisms by local receptor and
ligand based interactions. First, chemoaffinity was remodelled so that all
interactions occurred via signal transmission between the receptors expressed
on the growth cones of retinal ganglion cell axons and the ligands expressed
across the tectum. Signals induced either an attractive or repulsive
interaction to cause growth cones to ascend or descend the gradient of ligand
expression on the tectum.

We then explored the behaviour of models incorporating this chemoaffinity
mechanism alone and in conjunction with several competition mechanisms. We
found parameters for the best models of each type (chemo-only [G],
chemo-competition [GC], chemo-receptor-receptor [GI] and chemo-receptor-ligand
[GJ]) by an optimisation approach. We then compared each model's ability to
reproduce the surgical and genetic manipulation results in a manner similar to
that taken by \citet{simpson_simple_2011} to find the simplest model that
would fit the data.

\subsection*{Receptor and ligand expression}

Each retinal ganglion cell projects $n$ growth cone agents (referred to
as \emph{branches}) which carry a set of 4 receptors, $r_i$, and 4 ligands,
$l_i$, indexed by $i$ and expressed at levels determined by the cell's
location on the retina. The expression levels of the receptors and ligands
vary with respect to the cell's position on the retinal surface. Each receptor
or ligand varies only with respect to one dimension.  Receptor expression
gradients are arranged in orthogonal pairs with the gradient of $r_0$ being
orthogonal to that of $r_1$. $r_2$, whose gradient is opposite to $r_0$, is
orthogonal to $r_3$. Ligands are also arranged in orthogonal pairs, with $l_0$
opposing $r_0$ (as $r_0$ increases, $l_0$ decreases) and $l_1$ orthogonal to
$l_0$ and opposing $r_1$.

Because there is convincing evidence that EphA and EphB receptors are
expressed in exponentially increasing
patterns \citep{reber_relative_2004,feldheim_genetic_2000,brown_topographic_2000,koulakov_stochastic_2004},
we use an exponential form for retinal receptor expressions, in common with
other modelling
studies~\citep{reber_relative_2004,koulakov_stochastic_2004,simpson_simple_2011}.
We adopt the same precise form for the retinal receptor expression
as \citet{simpson_simple_2011} which is
\begin{equation} \label{e:retrcpt}
r_i(x,y) = 1.05 + 0.26 \exp(2.3 x)
\end{equation}

The simulated tectum expresses the same types of receptors and ligands, also
in orthogonal pairs of gradients. In this work, we did not consider the effect
of reverse signalling from tectal receptors to RGC ligands so we only modelled
tectal ligand expression, $L_i$. Although several studies model tectal ligand
expression with exponential functions \citep{koulakov_stochastic_2004}, the
experimental evidence for ligand expression is more ambiguous (Supp. Fig
\textbf{X}). Correspondingly, we leave open the possibility that tectal ligand
expression may be modelled by exponential, linear or logarithmic functions:
\begin{equation} \label{e:tecligexp}
L_i^{\text{exp}}(x,y) = 1.05 + 0.26 \exp(2.3 x)
\end{equation}
or
\begin{equation}
L_i^{\text{lin}}(x,y) = 1.31 + 2.33 x;
\end{equation}
or
\begin{equation}
L_i^{\text{log}}(x,y) = 2.32 + 1.29 \log (2.3(x+0.2))
\end{equation}
Each function has approximately the same expression values at $x=0$ and
$x=1$. Opposing functions are obtained by substituting $(1-x)$ for $x$.

While we do not explicitly name $r_0$, $r_1$,
etc.~as \emph{EphA}, \emph{EphB}, the suggestion is that $r$ includes EphA,
EphB, Ryk \citep{schmitt_wntryk_2006} and
Neogenin \citep{rajagopalan_neogenin_2004} receptors and that $l$ and $L$
include the ephrin-A, ephrin-B, Wnt3 \citep{schmitt_wntryk_2006} and
RGM \citep{monnier_rgm_2002} ligands, each of which has been shown to play a
role in retinotectal map formation.

Figure~\ref{f:ex} shows colour maps of receptor and ligand expression, in this
case tectal ligand expression maps are linear. Note that the expression is
discretized, and in simulations, the gradient is computed numerically from the
discretized expression. This makes it straightforward to perform tectal and
retinal `graft' manipulations. Coordinates within the simulated retina and
tectum are normalised, with 1 corresponding to 2 mm on the tectum and on the
retina for mouse \citep{reber_relative_2004}. This means that growth cones of
approximate radius 5 $\mu$m = 0.005 mm \citep{goodhill_can_2016} have a
radius $r$ = 0.0025 in the arbitrary units of the simulation. For a realistic
simulation based on mouse, the approximate RGC density in the retina is around
44000 cells per retina \citep{jeon_major_1998}. This sets the side length of
the retina to $\sqrt{44000} \approx 200$. This results in a simulation of
prodigious computational size, which although tractable (10 s per step,
requiring several hours, rather than days to run) is not convenient. To
maintain the relationship between growth cone size and cell density, we permit
a smaller number of growth cones ($400\times8=3200$) but keep the fractional
area of the branches at 0.86 ($44000 \times \pi r^2$) (if there is one cone
per axon) or 6.9 (if there are 8 cones per axon).

\subsection*{Movement}

At each timestep in a simulation, the position, $\mathbf{x}_{b,t}$, at time
$t$ of each branch $b$ is updated according to
%
\begin{equation}
\mathbf{x}_{b,t+1} = \mathbf{x}_{b,t} + \mathbf{M}_{b}
\end{equation}
%
where $\mathbf{M}_{b}$, the movement vector, is the sum of a chemoaffinity
effect, $\mathbf{G}_b$, axon-axon competition, $\mathbf{C}_b$, an axon-axon
receptor-receptor interaction, $\mathbf{I}_b$, and axon-axon receptor-ligand
interaction, $\mathbf{J}_b$, and a `border effect',
$\mathbf{B}_b$, which acts to retain branches within the tectal region:
%
\begin{equation}
\mathbf{M}_{b} = m_1 \mathbf{G}_b + m_2 \mathbf{C}_b + m_3 \mathbf{I}_b +
m_4 \mathbf{J}_b + m_5 \mathbf{B}_b
\end{equation}
%
evaluated at time $t$. $m_1$--$m_5$ are scalar parameters. We typically
considered models for which some of $m_2$ to $m_4$ were fixed to zero. A model
for which $m_3$ and $m_4$ were zero was named a `$\mathbf{GC}$' model as it
combined the chemoaffinity effect and space-based competition; a model for
which $m_2$ was zero was a `$\mathbf{GIJ}$' model and so on.

\subsection*{Chemoaffinity}

In this model, the signal transmitted when a ligand binds to a receptor on a
branch (forward signalling) can lead to one of two effects; the signal may
induce an attraction towards the ligand-expressing region or a repulsion away
from it.
%
For axon-tectum interactions, we implemented attraction as a climbing of the
tectal ligand expression gradient and repulsion as gradient descent.
%
To determine the strength of the effect we assumed a purely linear receptor
binding model, and set the chemotactic movement vector of the branch $b$ at
location $\mathbf{x}_b$ on the tectum to be
%
\begin{equation}
\mathbf{G}_b = \sum_i^N F_i\,r_{i,b} \nabla L_i(\mathbf{x}_b)
\end{equation}
%
where $r_{i,b}$ is the receptor expression on branch $b$ for ligand-receptor
pair $i$, $\nabla L_i$ is the gradient of expression of ligand $i$ on the
tectum and $F_i$ denotes the direction of the interaction induced when a
molecule of ligand $i$ binds to a receptor $i$ molecule. $F_i$ takes the value
-1 for a repulsive interaction or 1 for an attractive interaction.

\subsection*{Axon-axon interactions}

We investigated three forms of competition that could be combined with the
chemoaffinity effect. We investigated space-based competition
(`$\mathbf{C}$'), as described in \citet{simpson_simple_2011}, a
receptor-receptor signalling mechanism (`$\mathbf{I}$'), similar to that
described in \citet{simpson_simple_2011}, and a receptor-ligand mass-action
signalling model (`$\mathbf{J}$'). Our aim was to determine whether either of
the receptor/ligand based signalling mechanisms could replace the space-based
competition.

\subsection*{Space based (or self-other) competition}

Space based competition is a repulsive effect which acts between any two
branches, whether they originate from different retinal ganglion cells or from
the same cell. It assumes that there is some kind of self-other signal which
acts to discourage two cells from approaching each other.
%
\begin{equation}
\mathbf{C}_b = \frac{1}{|B_{b,c}|} \sum_k \hat{\mathbf{x}}_{kb} W
\end{equation}
where the distance based weight $W$ is given by
\begin{equation}
W_c(d_{kb}) = \begin{cases}
      1 - \frac{d_{kb}}{2r_c}   & d_{kb} \leq 2r_c \\
     0 & d_{kb} > 2r_c
     \end{cases}
\end{equation}
for two growth cones of interaction radius $r_c$ separated by a distance
$d_{kb}$. $B_{b,c}$ is the set of growth cones within a distance $2r_c$ of
branch $b$.

\subsection*{Receptor-receptor axon-axon interactions}

We modelled interactions between axon growth cones due to signalling between
receptors of the same type. For these interactions, repulsion caused a
movement of branch $b$ along a unit vector, $\hat{\mathbf{x}}_{kb}$, from
branch $k$ to branch $b$ causing them to move further apart; attraction caused
the opposite movement. This interaction, $\mathbf{I}_b$, acting on branch $b$,
is given by
%
\begin{equation}
\mathbf{I}_b
= \frac{1}{|B_{b,I}|} \sum_k \sum_i \hat{\mathbf{x}}_{kb}\,W_I \qquad \mathrm{if}~r_{i,b}
/ r_{i,k} < 1.1
\end{equation}
%
the weight here is defined as a function of $r_I$:
%
\begin{equation}
W_I(d_{kb}) = \begin{cases}
      1 - \frac{d_{kb}}{2r_I}   & d_{kb} \leq 2r_I \\
     0 & d_{kb} > 2r_I
     \end{cases}
\end{equation}
$r_{i,k}$ is the expression of receptor $i$ on branch $k$.

\subsection*{Receptor-ligand axon-axon interactions}

%
\begin{equation}
% No \frac{1}{|B_b|} in this one. Really?
\mathbf{J}_b = \sum_k \hat{\mathbf{x}}_{kb}\,Q_J(d_{kb})
\end{equation}
%
where $Q_J(d_{kb})$ is the signalling strength between two growth cones of
receptor-ligand interaction radius $r_j$ a distance $d_{kb}$ from one another,
given by
%
\begin{equation}
Q_J(d_{kb}) = \begin{cases}
     \sum_i^N F_i\,r_{i,b}\,l_{i,k}    & d_{kb} \leq 2r_j \\
     0 & d_{kb} > 2r_j
     \end{cases}
\end{equation}
%
The sign of $Q_J$ (which is dependent on the values of $F_i$) determines whether
the interaction, $\mathbf{I}_b$, is repulsive or attractive. $l_{i,k}$ is the
expression of ligand $i$ on branch $k$.

\subsection*{Border effect}

We implemented a border effect based on gradient following by assuming that
there is some other molecular signal which acts on all branches near the
boundary of the tectal tissue. For a branch with position $(x,y)$, $\mathbf{B}_b$ is
given by:
%
\begin{equation}
B_{b,x} = \begin{cases}
        r-x      & x<r \\
        1-r-x    & x>1-r
\end{cases}
B_{b,y} = \begin{cases}
        r-y      & y<r \\
        1-r-y    & y>1-r
\end{cases}
\end{equation}
%
Thus $\mathbf{B}_b$ is equivalent to the action of a repulsive signalling
molecule expressed around the border of the tectum, whose expression increases
quadratically outside the tectum and affects any branch touching (or outside)
the boundary

\subsection*{Initial conditions}

Branches were randomly distributed in a stripe at the rostral side of the
tectum at the start of each simulation. Each RGC axon was assigned a random
initial position coordinate:
\begin{equation}
\mathbf{x}_{\text{axon},t=0} = (U(0,1), U(-0.2,0))
\end{equation}
where U(p,q) is a number selected from a random uniform distribution in the
range $[p,q)$. Each of the $n$ branches per RGC axon was given its parent
axon's initial position, plus a randomly generated offset with coordinates
derived from a normal distribution of mean 0 and standard deviation 0.1.

\subsection*{Assumptions}

% All the extra bits of info required to specify the simulation
To enable a simulation of the model described above it is necessary to make
some assumptions about the nature of the receptor-ligand signalled
interactions (i.e. the values of $F_i$ that determine whether interactions are
attractive or repulsive) and about the pattern of ligand expression on the
tectal surface. Unless stated otherwise, we assumed that: \textbf{i)} all
receptor-ligand signalled interactions are repulsive ($F_i=-1, \forall i$), as
for EphA-ephrinA
coupling \citep{drescher_vitro_1995,nakamoto_topographically_1996};
and \textbf{ii)} all tectal ligand expressions vary exponentially along a
given axis, following Eq.\,\ref{e:tecligexp}.

In all simulations, the border effect, $B_b$, had the same magnitude with
$m_5=0.5$. The number of branches per axon was 4 unless otherwise stated.

@@

The first question we tried to answer was whether a gradient following
mechansim, acting alone, is sufficient to pattern the tectum. To do so, we
simulated the retina and tectum as unit square regions and defined retinal
receptor expression and tectal ligand expression across these squares
(Fig.\,\ref{f:ex}).

@@

\color{black}
Each discrete element in Fig.\,\ref{f:ex}B projects one RGC axon with 4
branches to the tectum, carrying the receptor expression at the retinal
element's location.
At each timestep in a simulation, the position, $\mathbf{x}_{b,t}$, at time
$t$ of each branch $b$ is updated according to
%
\begin{equation}
\mathbf{x}_{b,t+1} = \mathbf{x}_{b,t} + \mathbf{M}_{b}
\end{equation}
%
where $\mathbf{M}_{b}$, the movement vector, is the sum of a chemoaffinity
effect and a `border effect', $\mathbf{B}_b$, which acts to retain branches
within the tectal region:
%
\begin{equation}
\mathbf{M}_{b} = m_1 \mathbf{G}_b + 0.5 \mathbf{B}_b
\end{equation}
%
evaluated at time $t$.

% G - chemoaffinity description
The chemoaffinity component, $\mathbf{G}_b$, is governed by receptor-ligand
signalling. The signal transmitted when a tectal ligand binds to a RGC
receptor on a branch (forward signalling) can lead to one of two effects; the
signal may induce an attraction towards the ligand-expressing region or a
repulsion away from it.
%
For axon-tectum interactions, we implemented attraction as a climbing of the
tectal ligand expression gradient and repulsion as gradient descent.
%
To determine the strength of the effect we assumed a purely linear receptor
binding model, and set the chemotactic movement vector of the branch $b$ at
location $\mathbf{x}_b$ on the tectum to be
%
\begin{equation}
\mathbf{G}_b = \sum_i^N F_i\,r_{i,b} \nabla L_i(\mathbf{x}_b)
\end{equation}
%
where $r_{i,b}$ is the receptor expression on branch $b$ for ligand-receptor
pair $i$, $\nabla L_i$ is the gradient of expression of ligand $i$ on the
tectum and $F_i$ denotes the direction of the interaction induced when a
molecule of ligand $i$ binds to a receptor $i$ molecule. $F_i$ takes the value
-1 for a repulsive interaction or 1 for an attractive interaction.
%
We assumed that all
receptor-ligand signalled interactions are repulsive ($F_i=-1, \forall i$), as
for EphA-ephrinA
coupling \citep{drescher_vitro_1995,nakamoto_topographically_1996}.
%
$m_1$ is a
scalar parameter which controls how much movement is generated for a given
receptor-ligand signal size.
% end chemoaffinity description

% border effect description
The border effect is also based on gradient following by assuming that
there is some other molecular signal which acts on all branches near the
boundary of the tectal tissue. For a branch with position $(x,y)$, $\mathbf{B}_b$ is
given by:
%
\begin{equation}
B_{b,x} = \begin{cases}
        r-x      & x<r \\
        1-r-x    & x>1-r
\end{cases}
B_{b,y} = \begin{cases}
        r-y      & y<r \\
        1-r-y    & y>1-r
\end{cases}
\end{equation}
%
Thus $\mathbf{B}_b$ is equivalent to the action of a repulsive signalling
molecule expressed around the border of the tectum, whose expression increases
quadratically outside the tectum and affects any branch touching (or outside)
the boundary.
% end border effect description

% Initial conditions
Branches were initially randomly distributed in a stripe at the rostral side
of the tectum at the start of each simulation. Each RGC axon was assigned a
random initial position coordinate:
\begin{equation}
\mathbf{x}_{\text{axon},t=0} = (U(0,1), U(-0.2,0))
\end{equation}
where U(p,q) is a number selected from a random uniform distribution in the
range $[p,q)$. Each of the 4 branches per RGC axon was given its parent
axon's initial position, plus a randomly generated offset with coordinates
derived from a normal distribution of mean 0 and standard deviation 0.1.
% end initial conditions

@@

Fig\,\ref{f:chsurg} shows the results of 6 simulated surgical
manipulations. In row A, a patch of tectal tissue has been cut out and rotated
by a quarter turn, as in Fig.\,\ref{f:trot90}. From left to right, graphs show
i) the expected final layout of the tectum, according to the results of
experimental work, with the colour code giving the origin location of the
axon's soma on the retina, ii) The locations of individual axon branches at
simulation end, iii) the centroids of each axons' branches and iv) a rather
useless selected axons box \textbf{FIXME}. Simulation time was 1000 time steps. The final
locations of branches and axon centroids indicates that a qualitative rotation
of the graft patch has occurred, with red branches inside the patch being
located at the lateral-caudal corner rather than the medial-caudal corner. A
portion of the graft region (the medial side, after rotation) is free of
branches, indicating that the pattern is partial. A similar result is seen for
the 180\degree~example in row B, in whcih the caudal half of the rotated patch
is free of branches, but the rostral half contains branches with the correctly
rotated orientations.

@@

Previous work has shown that competition can contribute to the correct layout
of the retinotectal system \citep{stuff} and to the patterning of other
systems including the murine cortical whisker
barrels\citep{james_modelling_2020} \textbf{(FIXME owt else?)}.

@@

These were: space-based competition (`$\mathbf{C}$'), as described
in \citet{simpson_simple_2011}; a receptor-receptor signalling mechanism
(`$\mathbf{I}$'), similar to that described in \citet{simpson_simple_2011},
and a receptor-ligand mass-action signalling model
(`$\mathbf{J}$'). \citet{simpson_simple_2011} found that both a space based
competition and a receptor-receptor signalling mechanism were required to
reproduce the full range of surgical and genetic manipulations explored in
that work. The first of these, space-based competition, offers no mechanism by
which axon growth cones are able to communicate and the second mechanism
explored by \citet{simpson_simple_2011} was implemented only for a single axis
in the model (EphA receptors, expressed in a gradient along the XX retinal
axis). We wondered whether a single signalling mechanism could provide a
suitable competitive, repulsive effect. To investigate, we compared both of
the \citet{simpson_simple_2011} effects along with a third possible signalling
mechanism involving the same receptor-ligand interactions employed by the
gradient following mechansim, because the ligands expressed on the tectum are
also expressed by the retinal ganglion cells.

\subsection*{Space based (or self-other) competition}

@@

\subsection*{Receptor-receptor axon-axon interactions}

We modelled interactions between axon growth cones due to signalling between
receptors of the same type. For these interactions, repulsion caused a
movement of branch $b$ along a unit vector, $\hat{\mathbf{x}}_{kb}$, from
branch $k$ to branch $b$ causing them to move further apart; attraction caused
the opposite movement. This interaction, $\mathbf{I}_b$, acting on branch $b$,
is given by
%
\begin{equation}
\mathbf{I}_b
= \frac{1}{|B_{b,I}|} \sum_k \sum_i \hat{\mathbf{x}}_{kb}\,W_I \qquad \mathrm{if}~r_{i,b}
/ r_{i,k} < 1.1
\end{equation}
%
the weight here is defined as a function of $r_I$:
%
\begin{equation}
W_I(d_{kb}) = \begin{cases}
      1 - \frac{d_{kb}}{2r_I}   & d_{kb} \leq 2r_I \\
     0 & d_{kb} > 2r_I
     \end{cases}
\end{equation}
$r_{i,k}$ is the expression of receptor $i$ on branch $k$.

\subsection*{Receptor-ligand axon-axon interactions}

%
\begin{equation}
% No \frac{1}{|B_b|} in this one. Really?
\mathbf{J}_b = \sum_k \hat{\mathbf{x}}_{kb}\,Q_J(d_{kb})
\end{equation}
%
where $Q_J(d_{kb})$ is the signalling strength between two growth cones of
receptor-ligand interaction radius $r_j$ a distance $d_{kb}$ from one another,
given by
%
\begin{equation}
Q_J(d_{kb}) = \begin{cases}
     \sum_i^N F_i\,r_{i,b}\,l_{i,k}    & d_{kb} \leq 2r_j \\
     0 & d_{kb} > 2r_j
     \end{cases}
\end{equation}
%
The sign of $Q_J$ (which is dependent on the values of $F_i$) determines whether
the interaction, $\mathbf{I}_b$, is repulsive or attractive. $l_{i,k}$ is the
expression of ligand $i$ on branch $k$.

@@

\color{grey}
\section{ZEROTH DRAFT/NOTES}

\subsection*{Normal development}

%
Fig.\,\ref{f:wt}Ai shows a dual-colour map of retinal ganglion cell positions
on the square retina, with red encoding position on the $x$ axis (temporal to
nasal) and green encoding position on the $y$ axis (ventral to
dorsal). Fig.\,\ref{f:wt}Aii presents a fishnet plot of the expected location on
the tectal surface of each RGC at the end of the development process. Thus,
the `red' RGC from the ventro-nasal corner should map to the caudo-medial
corner of the tectum and the `green' RGC from the dorso-temporal corner should
find its way to the rostro-lateral corner on the tectum. The lines between the
dots of Fig.\,\ref{f:wt}Aii indicate the adjacency relationships between
cells. Thus, colour in each panel indicates a branch or axon's location of
origin on the retina.
%
Fig.\,\ref{f:wt}B shows the initial state of the system, with branches
stochastically arranged around the rostral edge of the tectum. Each dot
represents the centroid of $n=8$ branches per RGC axon. Figs.\,\ref{f:wt}C and \ref{f:wt}D
show the development of the system with time, with Fig.\,\ref{f:wt}D showing a
well ordered arrangement, with a majority of adjacency relationships matching
the experimental prediction (i.e. there are few crossed lines). The
intermediate timepoint at $t=40$ indicates that branches quickly cover the simulated
tectal surface, with final adjacency relationships developing over a longer
timescale. This dynamic behaviour is also evident in Fig.\,\ref{f:wt}E, which
shows \textbf{RMS error or some metric} of the axon centroids with respect to
the experimental prediction. Fig.\,\ref{f:wt}F shows the final
location of each branch in the simulation, which gives a more continuous view
of the order present in Fig.\,\ref{f:wt}D. Finally, Fig.\,\ref{f:wt}G shows the
branches and their centroid path history for five selected RGCs, with path
histories strongly resembling those presented by \citet{simpson_simple_2011}.

This result shows that a gradient-following model based on receptor-ligand
interactions (both for axon-to-tectum and axon-to-axon interactions) is
able to generate the basic mapping observed in the retinotectal projection.

\subsection*{Manipulations}

To determine if the model can explain the many surgical and genetic
manipulations described in the literature
\citep*[for a review, see][]{goodhill_retinotectal_1999} we performed
simulations matching those in \citet{simpson_simple_2011} in which we either
rotated or translocated grafts of the tectum; ablated portions of the retina
and/or the tectum; or manipulated the receptor or ligand expressions in the
model.

\subsubsection*{Surgical Manipulations}

Fig.\,\ref{f:surg1} shows the result of simulating three surgical
manipulations alongside the wildtype result, which is shown on Row A. Rows B
and C show graft rotation simulations (90$\degree$ and 180$\degree$). In each
case an 8x8 square on the tectum (each unit is 0.05x0.05) is rotated. The
rotational sense can be observed clearly in the `Branches' column; for the
90$\degree$ rotation, the top-right (caudal-lateral) corner of the rotated
square is red, whereas the top-left corner of the wildtype tectum is red. \textbf{etc}.

%\includegraphics[width=\linewidth]{./images/fig_manipulations.png}

\begin{figure}
\begin{subfigure}{\linewidth}
\includegraphics[width=\linewidth]{./images/j4_eE_GJ_wt_rotcmp.png}
\end{subfigure}
\begin{subfigure}{\linewidth}
\includegraphics[width=\linewidth]{./images/j4_eE_GJ_tecrot90.png}
\end{subfigure}
\begin{subfigure}{\linewidth}
\includegraphics[width=\linewidth]{./images/j4_eE_GJ_tecrot180.png}
\end{subfigure}
\begin{subfigure}{\linewidth}
\includegraphics[width=\linewidth]{./images/j4_eE_GJ_tecswap.png}
\end{subfigure}
\caption{Surgical manipulations. Row A: wildtype. Row B: 90 degree
rotation. Row C: 180 degree rotation. Row D: Graft swap.}
\label{f:surg1}
\end{figure}

Fig.\,\ref{f:surg2} shows retinal ablation, tectal ablation, mismatch and
compound eye experiments (Stuart: see Simpson and Goodhill Fig. 4).

>>>>

@@

The first question we tried to answer was whether a gradient following
mechanism, acting alone, is sufficient to pattern the tectum. To do so, we
simulated the retina and tectum as unit square regions and defined retinal
receptor expression and tectal ligand expression across these squares
(Fig.\,\ref{f:ex}).

@@

\color{black}
Each discrete element in Fig.\,\ref{f:ex}B projects one RGC axon with 4
branches to the tectum, carrying the receptor expression at the retinal
element's location.
At each timestep in a simulation, the position, $\mathbf{x}_{b,t}$, at time
$t$ of each branch $b$ is updated according to
%
\begin{equation}
\mathbf{x}_{b,t+1} = \mathbf{x}_{b,t} + \mathbf{M}_{b}
\end{equation}
%
where $\mathbf{M}_{b}$, the movement vector, is the sum of a chemoaffinity
effect and a `border effect', $\mathbf{B}_b$, which acts to retain branches
within the tectal region:
%
\begin{equation} \label{e:mv}
\mathbf{M}_{b} = m_{\!_G} \mathbf{G}_b + 0.5 \mathbf{B}_b
\end{equation}
%
evaluated at time $t$.

% G - chemoaffinity description
%\subsection{Chemoaffinity}
The chemoaffinity component, $\mathbf{G}_b$, is governed by receptor-ligand
signalling. The signal transmitted when a tectal ligand binds to a RGC
receptor on a branch (forward signalling) can lead to one of two effects; the
signal may induce an attraction towards the ligand-expressing region or a
repulsion away from it.
%
For axon-tectum interactions, we implemented attraction as a climbing of the
tectal ligand expression gradient and repulsion as gradient descent.
%
To determine the strength of the effect we assumed a purely linear receptor
binding model, and set the chemotactic movement vector of the branch $b$ at
location $\mathbf{x}_b$ on the tectum to be
%
\begin{equation}
\mathbf{G}_b = \sum_i^N F_i\,r_{i,b} \nabla L_i(\mathbf{x}_b)
\end{equation}
%
where $r_{i,b}$ is the receptor expression on branch $b$ for ligand-receptor
pair $i$, $\nabla L_i$ is the gradient of expression of ligand $i$ on the
tectum and $F_i$ denotes the direction of the interaction induced when a
molecule of ligand $i$ binds to a receptor $i$ molecule. $F_i$ takes the value
-1 for a repulsive interaction or 1 for an attractive interaction.
%
We assumed that all
receptor-ligand signalled interactions are repulsive ($F_i=-1, \forall i$), as
for EphA-ephrinA
coupling \citep{drescher_vitro_1995,nakamoto_topographically_1996}.
%
$m_{\!_G}$ is a
scalar parameter which controls how much movement is generated for a given
receptor-ligand signal size.
% end chemoaffinity description

% \subsection{Border effect} % border effect description
The border effect is also based on gradient following by assuming that
there is some other molecular signal which acts on all branches near the
boundary of the tectal tissue. For a branch with position $(x,y)$, $\mathbf{B}_b$ is
given by:
%
\begin{equation}
B_{b,x} = \begin{cases}
        r-x      & x<r \\
        1-r-x    & x>1-r
\end{cases}
B_{b,y} = \begin{cases}
        r-y      & y<r \\
        1-r-y    & y>1-r
\end{cases}
\end{equation}
%
Thus $\mathbf{B}_b$ is equivalent to the action of a repulsive signalling
molecule expressed around the border of the tectum, whose expression increases
quadratically outside the tectum and affects any branch touching (or outside)
the boundary.
% end border effect description

% \subsection*{Initial conditions}
Branches were initially randomly distributed in a stripe at the rostral side
of the tectum at the start of each simulation. Each RGC axon was assigned a
random initial position coordinate:
\begin{equation}
\mathbf{x}_{\text{axon},t=0} = (U(0,1), U(-0.2,0))
\end{equation}
where U(p,q) is a number selected from a random uniform distribution in the
range $[p,q)$. Each of the 4 branches per RGC axon was given its parent
axon's initial position, plus a randomly generated offset with coordinates
derived from a normal distribution of mean 0 and standard deviation 0.1.
% end initial conditions

@@

Fig\,\ref{f:chsurg} shows the results of 6 simulated surgical
manipulations. In row A, a patch of tectal tissue has been cut out and rotated
by a quarter turn, as in Fig.\,\ref{f:trot90}. From left to right, graphs show
i) the expected final layout of the tectum, according to the results of
experimental work, with the colour code giving the origin location of the
axon's soma on the retina, ii) The locations of individual axon branches at
simulation end, iii) the centroids of each axons' branches and iv) a rather
useless selected axons box \textbf{FIXME}. Simulation time was 1000 time steps. The final
locations of branches and axon centroids indicates that a qualitative rotation
of the graft patch has occurred, with red branches inside the patch being
located at the lateral-caudal corner rather than the medial-caudal corner. A
portion of the graft region (the medial side, after rotation) is free of
branches, indicating that the pattern is partial. A similar result is seen for
the 180\degree~example in row B, in which the caudal half of the rotated patch
is free of branches, but the rostral half contains branches with the correctly
rotated orientations.

@@

Previous work has shown that competition can contribute to the correct layout
of the retinotectal system \citep{stuff} and to the patterning of other
systems including the murine cortical whisker
barrels \citep{james_modelling_2020} \textbf{(FIXME owt else?)}.

@@

These were: space-based competition (`$\mathbf{C}$'), as described
in \citet{simpson_simple_2011}; a receptor-receptor signalling mechanism
(`$\mathbf{I}$'), similar to that described in \citet{simpson_simple_2011},
and a receptor-ligand mass-action signalling model
(`$\mathbf{J}$'). The movement vector (Eq.\,\ref{e:mv}) becomes
%
\begin{equation} \label{e:mv2}
\mathbf{M}_{b} = m_{\!_G} \mathbf{G}_b +  m_c \mathbf{C}_b +  m_{\!_I} \mathbf{I}_b +  m_{\!_J} \mathbf{J}_b + 0.5 \mathbf{B}_b
\end{equation}
%

Although two of these competitive mechanisms were investigated
by \citet{simpson_simple_2011}, the first, space-based competition, offers no
mechanism by which axon growth cones are able to communicate and the second
mechanism was implemented only for a single axis in the model (EphA receptors,
expressed in a gradient along the XX retinal axis). We wondered whether a
\emph{single} signalling mechanism working alongside the `untuned'
chemoaffinity with ligand expression given by Eq.\,\ref{e:tecligexp} could
successfully reproduce the wildtype and manipulation results. To investigate,
we studied each competitive mechanism alongside chemoaffinity in turn, setting
$m_{\!_I}=m_{\!_J}=0$ to give a `$\mathbf{GC}$' model, $m_c=m_{\!_J}=0$ to
give a `$\mathbf{GI}$' model or $m_c=m_{\!_I}=0$ to give a `$\mathbf{GJ}$'
model.

\subsection*{Chemoaffinity plus space-based competition (GC)}

@@

The model defined by the receptor and ligand expressions of Eqs.\,\ref{e:retrcpt}
\& \ref{e:tecligexp} along with Eq.\,\ref{e:mv2} with $m_{\!_I}=m_{\!_J}=0$ has three
free parameters; $m_{\!_G}$, $m_c$ and $r_c$. To find the best combination of
these parameters, we reasoned that evolution would optimize the arrangement of
the wildtype pattern. We performed a simulated annealing optimization to
minimize the error metric, $\epsilon$, for the wildtype pattern
(Fig.\,\ref{f:ch}E) at simulation time 1000. The parameters returned were $m_c$
= 0.09959, $m_{\!_G}$ = 0.003841 and $r_c$ =
0.39918. Fig.\,\ref{f:nonexistent_yet} shows a simulation with these parameters.

\subsection*{Receptor-receptor axon-axon interactions (GI)}

We modelled interactions between axon growth cones due to signalling between
receptors of the same type. For these interactions, repulsion caused a
movement of branch $b$ along a unit vector, $\hat{\mathbf{x}}_{kb}$, from
branch $k$ to branch $b$ causing them to move further apart; attraction caused
the opposite movement. This interaction, $\mathbf{I}_b$, acting on branch $b$,
is given by
%
\begin{equation}
\mathbf{I}_b
= \frac{1}{|B_{b,I}|} \sum_k \sum_i \hat{\mathbf{x}}_{kb}\,W_{\!_I} \qquad \mathrm{if}~r_{i,b}
/ r_{i,k} < 1.1
\end{equation}
%
the weight here is defined as a function of $r_{\!_I}$:
%
\begin{equation}
W_{\!_I}(d_{kb}) = \begin{cases}
      1 - \frac{d_{kb}}{2r_{\!_I}}   & d_{kb} \leq 2r_{\!_I} \\
     0 & d_{kb} > 2r_{\!_I}
     \end{cases}
\end{equation}
$r_{i,k}$ is the expression of receptor $i$ on branch $k$.

\subsection*{Receptor-ligand axon-axon interactions (GJ)}

%
\begin{equation}
% No \frac{1}{|B_b|} in this one. Really?
\mathbf{J}_b = \sum_k \hat{\mathbf{x}}_{kb}\,Q_{\!_J}(d_{kb})
\end{equation}
%
where $Q_{\!_J}(d_{kb})$ is the signalling strength between two growth cones of
receptor-ligand interaction radius $r_{\!_J}$ a distance $d_{kb}$ from one another,
given by
%
\begin{equation}
Q_{\!_J}(d_{kb}) = \begin{cases}
     \sum_i^N F_i\,r_{i,b}\,l_{i,k}    & d_{kb} \leq 2r_{\!_J} \\
     0 & d_{kb} > 2r_{\!_J}
     \end{cases}
\end{equation}
%
The sign of $Q_J$ (which is dependent on the values of $F_i$) determines whether
the interaction, $\mathbf{I}_b$, is repulsive or attractive. $l_{i,k}$ is the
expression of ligand $i$ on branch $k$.

@@
